---
title: "Moore-Capstone-Project"
format: html
editor: visual
---

Setting up R

```{r}
#Clearing R's environment
rm (list = ls())
```

```{r}
#Downloading necessary packages
library(tidyverse)
library(ggfortify)
library(here)
```

Reading in and looking at data

```{r}
#Reading in data
eastwood <- read.csv(here("Data", "Eastwood23_superb_climate_data.csv"))
```

```{r}
#Looking at variables and their types
glimpse(eastwood)
```

```{r}
#Opening data set in new window
View(eastwood)
```

Fixing variable types to set up for analysis

```{r}
#Changing "sex" to a factor variable
eastwood$Sex = as.factor(eastwood$Sex)
```

I will be investigating the relationship between telomere length and both rainfall and temperature. Therefore, the y-value in the following data exploration will be rTL (relative telomere length), and the x-values will be Rainfall.nestling (total rainfall during nestling phase) and Tmin.nestling (the mean daily minimum temperature during nestling phase).

Plotting data as a scatterplot to view each individual point and begin to investigate relationships.

```{r}
#Plotting rainfall and telomere length
ggplot(eastwood, aes(x = Rainfall.nestling, y = rTL)) + geom_point() + ylab("Telomere length") + xlab("Rainfall")
```

There is no clear relationship between rainfall and telomere length. There are 4 rainfall points that could be outliers (x = 87.4, 87.4, 94.6, 94.6), so I am going to plot data as a box-and-whisker plot to see if they show up there as well before deciding to remove them.

```{r}
#Plotting minimum temperature and telomere length
ggplot(eastwood, aes(x = Tmin.nestling, y = rTL)) + geom_point() + ylab("Telomere length") + xlab("Temperature")
```

There is no clear relationship between temperature and telomere length. I do not see any temperature outliers.

Plotting data as a box-and-whisker plot to clearly show outliers.

```{r}
#Plotting rainfall and telomere length
ggplot(eastwood, aes(x = Rainfall.nestling, y = rTL)) + geom_boxplot() + ylab("Telomere length") + xlab("Rainfall")
```

```{r}
#Plotting minimum temperature and telomere length
ggplot(eastwood, aes(x = Tmin.nestling, y = rTL)) + geom_boxplot() + ylab("Telomere length") + xlab("Temperature")
```

I do not see any outliers in either plot of this data, so I am not going to remove the temperature points that appeared to be outliers on the scatterplot.

Getting summary statistics of data.
```{r}
summary(eastwood)
```
The mean telomere length is 0.6615. The mean rainfall amount was 85.23. The mean minumum temperature was 10.546. This is consistent with where the means looked to be on the plots. 

Saving revised data frame for future use. 
```{r}
write.csv(eastwood, here("Data/Eastwood23_superb_climate_data_revised.csv"))
```

