---
title: "Moore-Data-Analysis-Linear-Regression"
format: html
editor: visual
---

# Question 1: Is there a relationship between rainfall amounts and telomere length?

I am looking for a relationship between rainfall and telomere length, which are both ratio data. Since there is only one predictor variable, I will use a linear regression.

## Setting up work space

```{r}
#Clearing R's environment
rm (list = ls())
```

```{r}
#Activating necessary packages
library(tidyverse)
library(ggfortify)
library(here)
```

```{r}
#Reading in data
eastwood <- read.csv(here("Data", "Eastwood23_superb_climate_data_revised.csv"))
```

## Predicting analysis

#### Making a scatterplot as initial depiction of data.

Making predictions off of this scatterplot can allow me to determine if there are any mistakes when I do my analysis. The summary statistics in the analysis should be similar to the ones estimated from the scatterplot.

```{r}
#Making a scatterplot
ggplot(eastwood, aes(Rainfall.nestling, rTL)) + geom_point()
```

The telomere lengths range between 0.25-1.20, and the rainfall amounts range between 0-95. The slope should be about (1.2-.25)/(95-0) = 0.01. The y-intercept appears to be around 0.25. I do not expect there to be a significant relationship between telomere length and rainfall amounts. Also, there does not appear to be a relationship based just upon a visual evaluation of the graph.

#### Making a histogram of telomere length to continue to visualize data

```{r}
ggplot(eastwood, aes(x = rTL)) + geom_histogram()
```

Looks to be relatively normally distributed -- no strong outliers or concerning shapes

#### Making a histogram of rainfall to continue to visualize data

```{r}
ggplot(eastwood, aes(x = Rainfall.nestling)) + geom_histogram()
```

Data is skewed right-- there appears to be zero inflation. I recognize that this may skew the data, but I will still continue with a linear regression because it is the most reasonable test out of the ones we have learned.

## Making linear regression

```{r}
#Creating model
model_rainfall <- lm(rTL ~ Rainfall.nestling, data = eastwood)
```

#### Checking assumptions

The residuals vs fitted plot should have no humps or valleys to assume that a linear model is appropriate for this data. In the Q-Q plot, the data should be linear to assume that the data is normally distributed. In the scale-location plot, there should be no pattern of points, so the line should be straight, to assume that there is constant variance. The residuals vs leverage plot detects points that have too much influence on the data set, such as outliers.

```{r}
#Checking assumptions
autoplot(model_rainfall, smooth.color = NA)
```

The Q-Q plot looks good, but the residuals vs fitted, scale-location, and residuals vs leverage plots are not as linear as they should be. There are 4 points that might be skewing this data, but the reason that I did not remove them before was because these outliers did not show up on a boxplot in my data exploration stage (seperate file). Despite this, I still think that the data fits the assumptions for a linear model well enough.

#### Printing ANOVA results

```{r}
#ANOVA test
anova(model_rainfall)
```

The F-statistic is 0.125 and the p-value is 0.7236, which does not indicate a significant relationship between rainfall amount and telomere length. The degrees of freedom + residuals + 1 should add up to the number of rows, which is the case because 1 + 341 + 1 = 343.

#### Summary

```{r}
#Viewing results in a summary table
summary(model_rainfall)
```

The adjusted R-squared is -0.002564. This value indicates that the variation in telomere length is not highly associated with rainfall amounts.

## Making new figure

```{r}
ggplot(eastwood, aes(Rainfall.nestling, rTL))+
     geom_point()+
     geom_smooth(method = "lm")+
     labs(x = "Rainfall during nestling phase", y = "Telomere length")+
     theme_bw()
```

## Results

Rainfall amount during the nestling phase for superb fairy wrens does not impact telomere length (F~1,341~ = 0.1253, p = 0.7236, adjusted R-squared = -0.002564). I had hypothesised that rainy conditions could have disturbed habitats and thus created stressors that shortened telomeres during developement, but it seems as though rain was not enough of a stressor to shorten telomeres. 